The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-04-13 Paper.Rmd
comments added as suggested by PCR

All branches of ecology study relationships among environmental and biological variables. However, ubiquitously used approaches to studying such relationships, based on correlation and regression, provide only a small slice of the complex information content of the relationships. Statistical tools do exist that provide a complete description of relationships, based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, but not in ecology. We here examine the potential for copula approaches to improve our understanding of ecology. We find that informative copula structure in dependences between variables is exceedingly common across all the environmental, species-trait, phenological, population, community, and ecosystem functioning dataset we considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more strongly related in their left compared to right tails, or vice versa. Copula structure and tail dependence can help reveal causal relationships between variables. We describe two mechanisms by which observed copula structure and tail dependence can arise in ecological data: a Moran-like effect whereby dependence structures between environmental variables are inherited by ecological variables; and asymmetric or nonlinear influences of environments on ecological variables, such as under Liebig's law of the minimum. We describe consequences of copula structure, including impacts on extinction risk and the stability through time of ecosystem services. By documenting the importance of a complete statistical description of dependence, advancing conceptual frameworks, and demonstrating powerful methods, we aim to elevate copula approaches to widespread use in ecology.

@@

The second mechanism we explored, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $a=0.2$, and 
$b=0.1, -0.1, 0.5, -0.5$ in different simulations. 
Negative values of $b$ correspond to overcompensating dynamics 
and positive values correspond to undercompensating dynamics; larger-magnitude $b$
mean slower return time of the model to equilibrium after a perturbation. 
The $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. 
We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represent environmental effects that occur only for negative 
values of the environmental variables, and $g_2$ represent environmental effects 
that occur only for positive values. Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

@@

The hypothesis that copula structure and tail dependence between environmental 
variables measured through time in different locations can produce similar 
structure in ecological variables influenced by the environmental variables was
largely confirmed, though similarities between environmental- and 
ecological-variable copula structure and tail dependencies can be reduced when the
intrinsic dynamics of the ecological variable is strongly nonlinear.
For simulations using AR(1) models (see section on methods for Q2), copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.

@@

Our model with environmental effects which differed in the
two tails could produce simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure, depending on the value of $b$ used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetric tail dependence and non-N copula structure were visually apparent.
We also got similar results from simulations with $b=\pm 0.5$ (Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) effects of the noise ($g_1$, resp., $g_2$) 
produced lower-tail (resp., upper-tail) dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
non-Normal copula was always best fit for $b= \pm 0.1$ and $\pm 0.5$, but their model 
AIC-wieght averaged tail dependence could not reflect the asymmetry for each case 
when that dependnece was at too much extreme end.
However, our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) focused on those extreme tips
and overcome the shortcomings of model selection approach which fit data based on maximum likelihood.

@@

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., average regional pest abundance computed from local pest abundances) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in Appendix \ref{SM-Consequence_TL_simulation} 
showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results demonstrate that, for many
ecological phenomena, complete understanding is impossible without 
characterizing dependence between variables using copulas.

@@

<!--Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Maybe mention possible future work: 
study copula structure of environmental variables (such as temperature and precip) that are known 
to be important in ecology, including how such copula structure may be changing and whether it 
has a geography of the kind studied in Walter et al.-->
One may expect our Moran-like mechanism for non-N copula structure and tail dependence to 
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures have often been found in prior analyses of 
environmental variables. If large meteorological events are also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made at different locations. Non-extreme values, not associated with extreme events, may
be more associated with local weather phenomena, and therefore 
less prone to being correlated across large areas.
Using a bivariate mixed model approach with copulas, 
@Serinaldi2008 examined the spatial dependence 
of rainfall observations (7 years with 30 minute time resolution) 
within a rain gauge network in Central Italy. Gumbel or Student 2-copulas 
were possible candidates for modelling bivariate dependence at several time scales, and 
neither of these is an N copula. @liu2016probabilistic used a copula 
based approach in assessing the drought risk with joint probability distribution 
between vegetation and climatic variables across China. 
In a majority of their cases, they found 
non-N copula structure in the dependence between a vegatation index 
and climatic variables (e.g., precipitation and temperature). 
Another long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a Gumbel copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables 
(duration and severity) in the study area. Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial data sets at different 
time scales [@Goswami2018].

@@

We thank the many contributors to the large datasets we used; D. Stevens and P. Verrier for data extraction; and Joel E. Cohen, Jonathan Walter, Thomas Anderson, and Lei Zhao for helpful suggestions. We thank James Bell of the Rothamsted Insect Survey (RIS). 
The Rothamsted Insect Survey, a National Capability, is funded by the Biotechnology and Biological Sciences Research Council under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National Science Foundation grants 0075516 and 1442595 and the James S McDonnell Foundation.

>>>>

All branches of ecology study relationships among environmental and biological variables. However, ubiquitously used approaches to studying such relationships, based on correlation and regression, provide only a small slice of the complex information content of the relationships. Other statistical tools exist that provide a complete description of relationships, based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, but not in ecology. We here examine the potential for copula approaches to improve our understanding of ecology. We find that informative copula structure in dependences between variables is exceedingly common across all the environmental, species-trait, phenological, population, community, and ecosystem functioning dataset we considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more strongly related in their left compared to right tails, or vice versa. Copula structure and tail dependence can help reveal causal relationships between variables. We describe two mechanisms by which observed copula structure and tail dependence can arise in ecological data: a Moran-like effect whereby dependence structures between environmental variables are inherited by ecological variables; and asymmetric or nonlinear influences of environments on ecological variables, such as under Liebig's law of the minimum. We describe consequences of copula structure, including impacts on extinction risk and the stability through time of ecosystem services. By documenting the importance of a complete statistical description of dependence, advancing conceptual frameworks, and demonstrating powerful methods, we aim to elevate copula approaches to widespread use in ecology.

@@

The second mechanism we explored, which was also summarized in the 
Introduction,
relates to Liebig's law of the minimum, or, more generally, to the fact that 
environmental variables can often influence ecological variables through nonlinear 
relationships. If an environmental variable influences an ecological variable
more strongly in one of its tails, we explored using a model 
whether the ecological variable
could then exhibit asymmetric tail dependence between measurements made in two
locations. Let $E_i(t)$ again be an ecological variable measured in locations
$i=1,2$. Assume $E_i(t)$ has dynamics that follow 
$E_i(t+1)=bE_i(t)+g(\epsilon_i(t))+a \delta_i(t)$, where $g$ is a function
describing how the environmental variable $\epsilon_i(t)$ influences
$E_i(t)$, and the $\delta_i(t)$ are standard-normally distributed 
local noise, independent across time and sampling locations. We used $a=0.2$, and 
$b=0.1, -0.1, 0.5, -0.5$ in different simulations. 
Negative values of $b$ correspond to overcompensating dynamics 
and positive values correspond to undercompensating dynamics; larger-magnitude $b$
means that the model has a slower return time to equilibrium after a perturbation. 
The $(\epsilon_1(t),\epsilon_2(t))$ were drawn, independently through
time, from a bivariate normal distribution with $\var(\epsilon_i)=1$ and 
$\cov(\epsilon_1,\epsilon_2)=0.8$. 
We used $g$ equal to $g_1$ or $g_2$ in two different simulations:
$g_1(\epsilon)$ equals $\epsilon$ if $\epsilon<0$ and $0$ otherwise; and 
$g_2(\epsilon)$ equals $\epsilon$ if $\epsilon>0$ and $0$ otherwise. Thus
$g_1$ represents environmental effects that occur only for negative 
values and $g_2$ only for positive values of the environmental variables.
Note that both $(\epsilon_1,\epsilon_2)$
and $(\delta_1,\delta_2)$ have N copula structure, being drawn from 
bivariate normal distributions. Hence the Moran-like mechanism discussed
previously cannot be operating in this model.

@@

The hypothesis that copula structure and tail dependence between environmental 
variables measured through time in different locations can produce similar 
structure in ecological variables influenced by the environmental variables was
largely confirmed, though similarities between environmental- and 
ecological-variable copula structure and tail dependencies can be reduced when the
intrinsic dynamics of the ecological variable are strongly nonlinear.
For simulations using AR(1) models (see section on methods for Q2), copula 
statistics were always very similar for noise and populations 
(Fig \ref{fig_Cause4copula_scatter_C_kend_ar1}, 
\ref{SM-fig_Cause4copula_C_kend_ar1}-\ref{SM-fig_Cause4copula_SJ_kend_ar1}),
though there were significant but minor differences for some statistics
and simulations.

@@

Our model with environmental effects which differed in the
two tails could produce simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure, depending on the value of $b$ used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetric tail dependence and non-N copula structure were visually apparent.
We also got similar results from simulations with $b=\pm 0.5$ (Figs \ref{SM-fig_ms_asym_b_0.5} - 
\ref{SM-fig_ms_asym_b_-0.5}).
Lower-tail (respectively, upper-tail) effects of the noise ($g_1$, respectively, $g_2$) 
produced lower-tail (respectively, upper-tail) dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) confirmed that
non-Normal copula was always best fit for $b= \pm 0.1$ and $\pm 0.5$, but their model 
AIC-wieght averaged tail dependence could not reflect the asymmetry for each case 
when that dependnece was at too much extreme end.
However, our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) focused on those extreme tips
and overcome the shortcomings of model selection approach which fit data based on maximum likelihood.

@@

<!--what we did and why is it important to study?-->
We have shown that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., average regional pest abundance computed from local pest abundances) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in Appendix \ref{SM-Consequence_TL_simulation} 
showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results demonstrate that, for many
ecological phenomena, complete understanding is impossible without 
characterizing dependence between variables using copulas.

@@

<!--Maybe a para on how likely one is to get copula structure via the Moran mechanism - considers
work on copula structure in environmental variables. Maybe mention possible future work: 
study copula structure of environmental variables (such as temperature and precip) that are known 
to be important in ecology, including how such copula structure may be changing and whether it 
has a geography of the kind studied in Walter et al.-->
One may expect our Moran-like mechanism for non-N copula structure and tail dependence to 
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures have often been found in prior analyses of 
environmental variables. A recent study [@agliardi2019relationship] found non-N 
copulas (mainly Gumbel and Survival Clayton) were better fit in modelling the 
upper tail dependence between extreme large values for global 
temperature and annual mean green house gas concentrations for an extensively
long datasets (1880-2017). If large meteorological events are also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made at different locations. Non-extreme values, not associated with extreme events, may
be more associated with local weather phenomena, and therefore 
less prone to being correlated across large areas.
Using a bivariate mixed model approach with copulas, 
@Serinaldi2008 examined the spatial dependence 
of rainfall observations (7 years with 30 minute time resolution) 
within a rain gauge network in Central Italy. Gumbel or Student 2-copulas 
were possible candidates for modelling bivariate dependence at several time scales, and 
neither of these is an N copula. @liu2016probabilistic used a copula 
based approach in assessing the drought risk with joint probability distribution 
between vegetation and climatic variables across China. 
In a majority of their cases, they found 
non-N copula structure in the dependence between a vegatation index 
and climatic variables (e.g., precipitation and temperature). 
Another long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a Gumbel copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables 
(duration and severity) in the study area. Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial data sets at different 
time scales [@Goswami2018].

@@

We thank the many contributors to the large datasets we used; D. Stevens and P. Verrier for data extraction; and Joel E. Cohen, Jonathan Walter, Thomas Anderson, and Lei Zhao for helpful suggestions. We thank James Bell of the Rothamsted Insect Survey (RIS). 
This survey, a UK Capability, is funded by the Biotechnology and Biological Sciences Research Council under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National Science Foundation grants 0075516 and 1442595 and the James S McDonnell Foundation.
